A universal of human social cognition: Children from 17 communities process gaze in similar ways

### 2024-10-16 paper/paper.Rmd
long version first draft

---
title             : "A universal of human social cognition: Children from 17 communities process gaze in similar ways"
shorttitle        : "gaze following across 17 communities"

@@

data%>%
  mutate(side = ifelse(targetcenterx < 960, "left", "right"),
         error = ifelse(clickdistfromtargetcenterx < 0, "left", "right"))%>%
  group_by(side)%>%
  summarise(left_bias = mean(error == "left"),
            right_bias = mean(error == "right"))

@@

Human socio-cognitive skills enable unique forms of communication and cooperation that provide a bedrock for cumulative culture and the formation of complex societies [@tomasello2020adaptive; @laland2021understanding; @wellman2014making; @henrich2016secret; @tomasello2003makes; @legare2019development; @heyes2018cognitive]. The eyes are the proverbial "window to the mind" and eye gaze is essential for many social reasoning processes [@shepherd2010following; @doherty2006development; @emery2000eyes; @frischen2007gaze]. Others’ eye gaze is used to infer their focus of visual attention, which is a critical aspect of coordinated activities, including communication and cooperation [@tomasello2007reliance; @scaife1975capacity; @langton2000eyes; @rossano2012gaze; @richardson2005looking; @sebanz2006joint]. During ontogeny, gaze following is an important aspect of many critical learning objectives that enable children to become functioning members of the society they grow up in, including language, social learning and joint action [@brownell2011early; @moore2008development; @brooks2005development; @carpenter1998social; @mundy2007attention; @stephenson2021gaze]. Because of the central role gaze following plays during human ontogeny, it has been widely argued that gaze following has been a target of natural selection [@tomasello2007reliance; @kano2023evolution; @emery2000eyes; @clark2023white]. This implies that the process by which humans use gaze direction to infer the focus of attention is universal. In this paper, we report a comprehensive cross-cultural study on the ontogeny of gaze following in which we shed light on the universal aspects of gaze following as well as sources of variation and their origins.

@@

The ability to follow gaze emerges early in development [@tang2023slow; @gredeback2010development; @byers2021development; @del2019developmental]. The earliest signs of gaze following have been found in infants as young as four months [@astor2021gaze; @d1997demonstration]. Throughout the first two years of life, children refine their abilities: they interpret gaze in mentalistic terms [@butterworth1991minds; @deak2000effects], for example, they follow gaze to locations outside their own visual field by moving around barriers [@moll200412]. Initially, children rely more on head direction than actual gaze direction [@michel2021effects]. In fact, when head and gaze direction diverge, children fail to accurately locate the agent's focus of attention up until 19 months of age [@lempers1979young].

From an evolutionary perspective, while many species are able to follow gaze based on head directions, uniquely human forms of joint action and communication require a more precise localization of other's attention and thus critically rely on gaze direction inferred from eye movements [@emery2000eyes; @hessels2020does]. In a recent study, @prein2024variation studied the development of gaze following based on eye movements across from three years up until old age. They found particularly steep developmental improvements in the preschool years resulting in a relatively stable level of accuracy from ten years onward and a slight decrease starting around age 40.

The studies reported thus far, all relied on data collected in western affluent settings. Such settings represent only a minority of the worlds population and are thus insufficient to make claims about universal aspects of human cognition [@nielsen2017persistent; @norenzayan2005psychological; @amir2020cross]. Three studies with infants and children from traditionally underrepresented parts of the world (Bhutan, India, Peru, Vanuatu) find that children start gaze following (including head direction) at similar ages, the rates of gaze following, however, differed between cultural settings [@hernik2019infant; @astor2022maternal; @callaghan2011early].

Rates and accuracy of gaze following do not just differ between cultural settings; there is also substantial variation within settings. In fact, the pivotal role of gaze following in many uniquely human activities has been studied by relating individual differences in gaze following to other phenomena -- both cross-sectionally and longitudinally [@carpenter1998social; @brooks2015connecting]. For example, gaze following at 10 months predicts language scores at 18 months of age [@brooks2005development; see also @macdonald2013eye]. Furthermore, difficulties with gaze following have been linked to developmental disorders, including Autism [@itier2009neural; @thorup2016altered; @thorup2018reduced] and -- at least in some cultural contexts -- to maternal postpartum depression [astor2022maternal]. Individual differences are also key to explaining the -- mainly social-interactional -- driving forces behind the development of gaze following. For example, it has been found that early attachment quality or the use of gaze in communicative interactions predict later rates of gaze following [@astor2020social; @movellan2002development; @senju2015early].

@@

The existence of variation in gaze following both within and between cultural settings raises the question of how to square these findings with the suggestion that gaze following is a fundamental building block of human social cognition and interaction and that eye movements are processed the same way in humans all over the world. Looking at other aspects of social cognition, one could easily make the argument that variation is the norm rather than the exception [see e.g., @miller2018contributions; @wellman2014making; @mayer2013synchrony; @dixson2018scaling; @taumoepeau2019cross]. As a first step, answering this question requires data from many different cultural settings. In a second step, however, we need to a way of detecting universal processes in such data. The traditional approach is to compare some sort of aggregate measure (mean level of performance, average age of onset) across cultural settings. Absolute differences in mean performance across communities are interpreted as a signal of different underlying cognitive processes [see e.g., @house2020universal; @van2018development; @blake2015ontogeny; @kanngiesser2022children]. No differences seemingly support the existence of a psychological universal. Such an approach, however, neglects the existence of within-cultural variation altogether [see also @gurven2018broadening].

In the present study, we want to take a different approach for which we assume that universal processes and variation can co-exist [@kline2018variation; @greenfield2003cultural; @jensen2012bridging]. Instead of starting with the outcome (performance in the task), we start with the process that generates the outcome. By defining this process, we make a proposal for the universal aspect of the process. At the same time, we define variable aspects in the process that generate individual differences. This allows us to define signatures that the process leaves behind that can be detected independent of absolute levels of performance.

For gaze following, we can use the computational model proposed by @prein2024variation to derive such predictions. They formalized the widely-held view that gaze following involves estimating a vector emanating from the eye center through the pupil [@todorovic2006geometrical; @yaniv1990heuristics; @butterworth1991minds; @symons2004you]. The key innovation of the model is that it explains how individuals may use the same  process but still differ in their measured abilities. The process always involves estimating a vector but also involves a degree of uncertainty because the eye center is not directly observable. Individuals are assumed to differ in their level of uncertainty with which they estimate the vector which causes differences in their observable behavior. Importantly, the assumed process leaves a key signature in the data that is observable independent of the absolute level of performance. In the present study, we therefore focus on this signature instead of absolute levels of performance when evaluating the claim whether there is evidence for a universal cognitive process underlying gaze following.

@@

The present study had three goals. First, to collect a comprehensive data set and study the ontogeny of gaze following beyond infancy across cultures. To make this possible, we used a semi-standardized task that required minimal assistance from an experimenter and no behavioral coding. The task is an animated picture book presented on a tablet screen. Children watched a balloon disappear behind a hedge. An agent followed the trajectory of the balloon with their eyes (Figure \@ref(fig:fig1)B). The key dependent variable was (im)precision, that is, the deviation between where the agent looked (where the balloon was) and the child’s response. The task's flexible implementation as a browser-based web-app allowed us to quickly tailor its visual and audio content to each cultural setting (visuals and audio). The task has been psychometrically evaluated and has shown to yield reliable individual-level measurements across communities and across ages [@prein2024tango; @prein2024measuring].

We collected data in 17 different communities across 14 countries and five continents. Communities covered a broad spectrum of geographical locations, social and political systems, languages, and subsistence styles. This diversity allowed us to overcome a common pitfall of cross-cultural studies that compare urban communities from the Global North to rural communities from the Global South [@barrett2020towards]. We aimed for large sample sizes within each community to contrast within- and between cultural variation. Our expectation regarding the first goal was to see substantial variation across cultures but even more variation between individuals. In all communities, we expected performance to increase with age.

The second goals was to look for signatures in the data of the universal gaze following process predicted by the model of @prein2024measuring. In the task, the hidden object is located on a horizontal plane at the lower end of the screen. The agent is located in the upper center of the screen (see Figure \@ref(fig:fig1)B). The model predicts trials in which the object is hidden further away from the center to be more difficult resulting in higher imprecision. The signature is thus a u-shaped relation between object location and imprecision (Figure \@ref(fig:fig12)).

Finally, we sought to explain individual differences in gaze following precision by linking them to methodological aspects of the study as well as aggregate measures of children's everyday social experience. Experience with tablets and touch screens co-varied with community and we expected children more familiar with this medium to perform better. Previous work suggested that gaze following is refined in social interaction [@movellan2002development; @senju2015early]. To approximate social interaction, we asked parents to fill out a questionnaire about household size and composition. We acknowledge that this measure approximates opportunities for social interaction in a rather crude way but we nevertheless expected children living in larger households and with more siblings (relative to their community) to be more accurate when following gaze.

@@

Visualization of the data also showed the predicted u-shaped pattern in all communities (see Fig. \@ref(fig:fig2)B). The visualization also showed that the u is not symmetric. Instead, imprecision seemed to be higher when the target lands on the left side of the screen.

@@

Following and understanding gaze is a foundational building block of human social cognition [@tomasello2007reliance; @scaife1975capacity; @langton2000eyes; @rossano2012gaze; @richardson2005looking; @sebanz2006joint]. A substantial body of work has explored the developmental onset of gaze following in a few selected cultural communities [@moore2008development; @tang2023slow; @gredeback2010development; @byers2021development]. The data reported here provides strong evidence that children from a large and diverse set of communities process others’ gaze in similar ways. We found key performance signatures predicted by a model treating gaze following as a form of social vector estimation process across all 17 communities. With the focus on individual-level processing signatures, the study goes beyond previous studies on gaze following -- focused on the onset of gaze following in infancy [@hernik2019infant; @callaghan2011early] -- as well as comprehensive cross-cultural studies that compared average developmental trajectories [@house2020universal; @van2018development; @blake2015ontogeny; @kanngiesser2022children].

Cultural settings are not homogeneous clusters but greatly overlap.

alternative models not really alternative processes - motor noise model predicts no u-shaped pattern.

The effect of touchscreen exposure wuggests that familiarity with the device used for data collection explains variation between communities. Children with more touchscreen experience were probably better at task handling and thus more likely to precisely touch the location they inferred the agent to look at. However, there was substantial variation between individuals that could not be explained by differential exposures to touchscreens alone. For example, in Malatya (Türkiye) where 100% of children had access to touchscreens there was still substantial variation between individuals (see Fig.\@ref(fig:fig1)B). This strongly indicates that other factors likely contributed to individual differences.

Social interaction has been highlighted as an important driver of social-cognitive development [e.g., @tomasello2019becoming; @carpendale2020makes; @barresi1996intentional; @perner1994theory; @rakoczy2022foundations] and thus we hypothesized (and pre-registered) that more opportunities for social interaction -- approximated by living in larger households with more children -- would be associated with higher levels of precision. When predicting performance by relative opportunities for social interactions within a community -- while accounting for absolute differences and the prevalence of touchscreens -- we found no strong associations between any of the demographic indicators and performance (see Supplementary Material).

The cognitive processes underlying gaze following might be rooted in humans’ evolved cognitive architecture, which is -- presumably -- later refined during social interaction [@astor2020social; @movellan2002development; @senju2015early]. The phylogenetic roots of these processes might possibly lie much deeper as primates from a wide range of species follow gaze [@rosati2009looking; @tomasello1998five; @itakura2004gaze; @kano2014cross]. Yet, similarities in overt behavior do not imply the same underlying cognitive processes. The present study defines clear performance signatures that can be explored in other species to test such evolutionary hypotheses.

Our study combined precise individual-level cognitive measurement and individual-level assessment of experience (here: touchscreen exposure) in a large and diverse sample to directly investigate the impact of specific cultural experiences on developmental outcomes. Instead of establishing universality by maximizing the cultural distance between two or three tested communities [@norenzayan2005psychological], this large-scale cross-cultural approach treats children’s cultural experience at scale, shedding light on the big "middle ground" of children’s cultural experience [@barrett2020towards].

The study has important limitations. The fact that performance in the task was correlated with exposure to touchscreens might have overshadowed other sources of variation. However, we think it is in an important innovation that we were able to account for this effect. Most developmental cross-cultural studies do not even question the portability of their measurement instruments. Importantly, the key result that the processing signatures were seen in all communities, is immune to this finding. The potential that lies in the otherwise precise individual-level measurement that the task achieves is largely unexploited. The questionnaire items only offer a very coarse picture into children's actual lived experiences. Whilst household size was a useful proxy for regular social interaction opportunities, the measure does not directly measure the factors that previous work has suggested to be related to the development of gaze following in younger children, such as attachment quality or the use of gaze in early communicative interactions [@astor2020social; @movellan2002development; @senju2015early]. Future work could increase the resolution with which everyday experiences in children from diverse communities are recorded to compare the drivers behind social-cognitive development as we observe it. Recent work in the field of language acquisition has shown how technological innovations allowed for direct recording of social interactions across communities which can be used to close this explanatory gap [@bergelson2023everyday; @donnelly2021longitudinal].

In sum, our work pioneers an approach that introduces computational modeling and precise individual-level measurement to the cross-cultural study of cognitive development. This approach allowed us to test for universals in the human cognitive architecture rather than just overt behavior. As such, it can serve as a blueprint for future research on a broad spectrum of cognitive abilities and offers a much-needed empirical foundation for theories on the nature of the human mind. Children from diverse communities deploy similar cognitive processes in interpreting gaze, pointing to a universal foundation of basic social cognition, which is refined during development.

>>>>

---
title             : "A universal of human social cognition: Children from 17 communities process gaze in similar ways"
shorttitle        : "Gaze following across 17 communities"

@@

@@

Human socio-cognitive skills enable unique forms of communication and cooperation that provide a bedrock for cumulative culture and the formation of complex societies [@tomasello2020adaptive; @laland2021understanding; @wellman2014making; @henrich2016secret; @tomasello2003makes; @legare2019development; @heyes2018cognitive]. The eyes are the proverbial "window to the mind" and eye gaze is essential for many social reasoning processes [@shepherd2010following; @doherty2006development; @emery2000eyes; @frischen2007gaze]. Others’ eye gaze is used to infer their focus of visual attention, which is a critical aspect of coordinated activities, including communication and cooperation [@tomasello2007reliance; @scaife1975capacity; @langton2000eyes; @rossano2012gaze; @richardson2005looking; @sebanz2006joint]. During ontogeny, gaze following is an important aspect of many critical learning objectives that enable children to become functioning members of the society they grow up in, including language, social learning and cooperation [@brownell2011early; @moore2008development; @brooks2005development; @carpenter1998social; @mundy2007attention; @stephenson2021gaze]. Because of the central role gaze following plays during human ontogeny, it has been widely argued that gaze following has been a target of natural selection [@tomasello2007reliance; @kano2023evolution; @emery2000eyes; @clark2023white]. This implies that the process by which humans use gaze direction to infer the focus of attention is universal. In this paper, we report a comprehensive cross-cultural study on the ontogeny of gaze following in which we shed light on the universal aspects of gaze following as well as sources of variation and their origins.

@@

The ability to follow gaze emerges early in development [@tang2023slow; @gredeback2010development; @byers2021development; @del2019developmental]. The earliest signs of gaze following have been found in infants as young as four months [@astor2021gaze; @d1997demonstration]. Throughout the first two years of life, children refine their abilities: they begin to interpret gaze in mentalistic terms [@butterworth1991minds; @deak2000effects], for example, they follow gaze to locations outside their own visual field by moving around barriers [@moll200412]. Initially, children rely more on head direction than actual gaze direction [@michel2021effects]. When head and gaze direction diverge, children often fail to accurately locate the agent's focus of attention up until 19 months of age [@lempers1979young].

From an evolutionary perspective, while many species are able to follow gaze based on head directions, uniquely human forms of joint action and communication require a more precise localization of other's attention and thus critically rely on gaze direction inferred from eye movements [@emery2000eyes; @hessels2020does]. In a recent study, @prein2024variation studied the development of gaze following based on eye movements from three years up until old age. They found particularly steep developmental improvements in the preschool years resulting in a relatively stable level of accuracy from ten years onward and a slight decrease starting around age 40.

The studies reported thus far, relied on data collected in western affluent settings. Such settings represent only a minority of the worlds population and are thus insufficient to make claims about universal aspects of human cognition [@nielsen2017persistent; @norenzayan2005psychological; @amir2020cross; @barrett2020towards]. Three studies with infants and children from traditionally underrepresented parts of the world (Bhutan, India, Peru, Vanuatu) find that even though children start gaze following (gaze and head direction combined) at similar ages, the rates of gaze following differ between cultural settings [@hernik2019infant; @astor2022maternal; @callaghan2011early].

Rates and accuracy of gaze following do not just differ between cultural settings; there is also substantial variation within settings. In fact, the pivotal role of gaze following in many uniquely human activities has been studied by relating individual differences in gaze following to other phenomena -- both cross-sectionally and longitudinally [@carpenter1998social; @brooks2015connecting]. For example, gaze following at 10 months predicts language scores at 18 months of age [@brooks2005development; @macdonald2013eye]. Furthermore, difficulties with gaze following have been linked to developmental disorders, including Autism [@itier2009neural; @thorup2016altered; @thorup2018reduced] and -- at least in some cultural contexts -- to maternal postpartum depression [astor2022maternal]. Individual differences are also key to explaining the -- mainly social-interactional -- driving forces behind the development of gaze following. For example, it has been found that early attachment quality or the use of gaze in communicative interactions predict later rates of gaze following [@astor2020social; @movellan2002development; @senju2015early].

@@

The existence of variation in gaze following both within and between cultural settings raises the question of how to square these findings with the suggestion that gaze following is a fundamental building block of human social cognition and interaction and that eye movements are processed the same way in humans all over the world. Looking at other aspects of social cognition, one could easily make the argument that variation is the norm rather than the exception [see e.g., @miller2018contributions; @wellman2014making; @mayer2013synchrony; @dixson2018scaling; @taumoepeau2019cross]. As a first step, answering this question requires data from many different cultural settings. In a second step, however, we need a way of detecting universal processes in such data. The traditional approach is to compare some sort of aggregate measure (mean level of performance, average age of onset) across cultural settings. Absolute differences between communities are interpreted as a signal of different underlying cognitive processes while no differences are taken to support the existence of a psychological universal [@house2020universal; @van2018development; @blake2015ontogeny; @kanngiesser2022children]. Such an approach, however, neglects the existence of within-cultural variation altogether [see also @gurven2018broadening].

In the present study, we want to take a different approach for which we assume that universal processes and variation can co-exist [@kline2018variation; @greenfield2003cultural; @jensen2012bridging]. Instead of starting with the outcome (performance in the task), we start with the process that generates the outcome. By defining this process, we make a proposal for the universal aspects of the process. At the same time, we define variable aspects that generate individual differences. This allows us to define signatures that the process leaves behind that can be detected independent of absolute levels of performance.

For gaze following, we can use the computational model proposed by @prein2024variation to derive such predictions. They formalized the widely-held view that gaze following involves estimating some sort of vector emanating from the eye center through the pupil [@todorovic2006geometrical; @yaniv1990heuristics; @butterworth1991minds; @symons2004you]. The key innovation of the model is that it explains how individuals may use the same process but still differ in their measured abilities. The process always involves estimating a vector but also involves a degree of uncertainty because the eye center is not directly observable. Individuals are assumed to differ in their level of uncertainty with which they estimate the vector which causes differences in their observable behavior. Importantly, the general vector estimation process leaves a key signature in the data that is detectable independent of the absolute level of performance. In the present study, we therefore focus on this signature instead of absolute levels of performance when evaluating the claim whether there is evidence for a universal cognitive process underlying gaze following.

@@

The present study had three goals. First, to collect a comprehensive data set and study the ontogeny of gaze following beyond infancy across cultures. To make this possible, we used a semi-standardized task that required minimal assistance from an experimenter and no behavioral coding. The task is an animated picture book presented on a tablet screen. Children watched a balloon disappear behind a hedge. An agent followed the trajectory of the balloon with their eyes (Figure \@ref(fig:fig1)B). The key dependent variable was (im)precision, that is, the deviation between where the agent looked (where the balloon was) and the child’s response. The task's flexible implementation as a browser-based web-app allowed us to quickly tailor its visual and audio content to each cultural setting (visuals and audio). The task has been psychometrically evaluated and has shown to yield reliable individual-level measurements across communities and ages [@prein2024tango; @prein2024measuring].

We collected data in 17 different communities across 14 countries and five continents. Communities covered a broad spectrum of geographical locations, social and political systems, languages, and subsistence styles. This diversity allowed us to overcome a common pitfall of cross-cultural studies that compare urban communities from the Global North to rural communities from the Global South [@barrett2020towards]. We aimed for large sample sizes within each community to contrast within- and between cultural variation. Our expectation regarding the first goal was to see substantial variation across cultures but even more variation between individuals. In all communities, we expected performance to improve with age.

The second goals was to look for signatures in the data of the universal gaze following process predicted by the model of @prein2024measuring. In the task, the hidden object lands on a horizontal plane at the lower end of the screen. The agent is located in the upper center of the screen (see Figure \@ref(fig:fig1)B). The model predicts that trials in which the object is hidden further away from the center to be more difficult, resulting in higher imprecision. The signature is thus a u-shaped relation between object location and imprecision (Figure \@ref(fig:fig12)).

Finally, we sought to explain individual differences in gaze following precision by linking them to methodological aspects of the study as well as aggregate measures of children's everyday social experience. Experience with tablets and touch screens co-varied with community and we expected children more familiar with this medium to perform better. Previous work suggested that gaze following is refined in social interaction [@movellan2002development; @senju2015early]. To approximate social interaction, we asked parents to fill out a questionnaire about household size and composition. We acknowledge that this measure approximates opportunities for social interaction in a rather coarse way but we nevertheless expected children living in larger households and with more siblings (relative to their community) to be more accurate when following gaze.

@@

```{r}
side_bias <- data%>%
  mutate(side = ifelse(targetcenterx < 960, "left", "right"),
         error = ifelse(clickdistfromtargetcenterx < 0, "left", "right"))%>%
  group_by(side)%>%
  summarise(left_bias = mean(error == "left"),
            right_bias = mean(error == "right"))
```

Visualization of the data also showed the predicted u-shaped pattern in all communities (see Fig. \@ref(fig:fig2)B). The visualization also showed that the u is not symmetric. Instead, imprecision was higher when the target landed on the left side of the screen.

@@

Following and understanding gaze is a foundational building block of human social cognition [@tomasello2007reliance; @scaife1975capacity; @langton2000eyes; @rossano2012gaze; @richardson2005looking; @sebanz2006joint]. A substantial body of work has explored the developmental onset of gaze following in early infancy and in a few selected cultural communities [@moore2008development; @tang2023slow; @gredeback2010development; @byers2021development]. The study reported here presents comparable data (i.e. collected via the same task with minimal superficial adjustments) on the development of gaze following in young children from 17 communities from five continents. We found substantial variation between cultural settings, both in average levels as well as the steepness of developmental trajectories. In that, however, individual-level variation greatly outweighed community-level variation. Despite this variation, we found evidence that the basic process of gaze following is the same across communities: we found key performance signatures predicted by a model conceptualizing gaze following as a form of social vector estimation in all communities. Individual-differences in gaze following were related to children's exposure to touchscreens but not to aggregate measures of opportunities for social interaction (i.e. household composition). This study provides evidence for a putative universal in basic social cognition and presents a new approach to studying cognitive processing in light of cross-cultural and individual variation.

The task we used has good individual-level measurement properties across cultures [@prein2024measuring]. This puts us in the position to contrast individual-level variability with community-level variability instead of dismissing the former as noise. The obvious pattern here was that cultural settings did not form homogeneous clusters but greatly overlapping distributions. That is, variation in the average developmental trajectories was small compared to variation between-individuals, both within communities and across them. Some individuals from the community with the highest average level of imprecision outperformed individuals from the community with the lowest average. Thus, the explanatory power of of community-level variables might be limited.

On the face of it, a community-level perspective on the results points to a urban vs. rural divide in the data. Importantly, our sample included urban settings from the Global North and South so that geographic location and living conditions were only partly confounded (we did not collect data in rural settings in the Global North). Case in point is Namibia, where we collected data in both types of settings with results mirroring the overall pattern. In previous work, such differences were often attributed to specific community-level differences in everyday experience that come with urbanization [e.g., @amir2020developmental; @mavridis2020development]. However, correlations identified this way remain speculative because urban and rural settings differ in a myriad of ways. Furthermore, this approach neglects within-community variation in everyday experience [@bohn2024understanding]. It is often chosen because the measures used are not suited to reliably quantify individual-level variation. Given the good measurement properties of our task and the individual-level assessment via the parental questionnaire, we were able to directly link aspects of experience and cognitive development, omitting the intermediate community-level.

We investigated both methodological and household composition as potential predictors. Familiarity with the device used for data collection explained variation between communities. Children with more touchscreen experience were probably better at task handling and thus more likely to precisely touch the location they inferred the agent to look at. Importantly, the model comparison showed that this relation did not vary substantially across communities. The effect, however, did not explain all variation between individuals. For example, in Malatya (Türkiye) where 100% of children had access to touchscreens there was still substantial variation between individuals. This strongly indicates that other factors likely contributed to individual differences.

Social-interactional variables have been linked to the development of gaze following in previous work [@astor2020social; @movellan2002development; @senju2015early]. Consequently, we predicted that opportunities for social interaction -- approximated by household size and composition -- would be linked to performance in the task while accounting for absolute differences and the prevalence of touchscreens. This was not the case. Yet, this result does not provide strong evidence for the absence of a relation between social-interactional variables and the development of gaze following, instead, we think it suggests that a more fine-grained approach is necessary to identify the relevant aspects of social interaction.

Despite substantial variation, we found the expected processing signatures in the data from all communities. Alternative accounts for how this pattern might have arisen did not explain the data well. However, these alternative approaches did not present viable alternative theoretical accounts for how participants followed gaze and why they differed from one another because they assumed that participants ignore gaze altogether. An alternative account involving the use of gaze cues would be that participants do not differ in the precision with which they estimate the gaze vector but that they differ only in the precision with which they touch the inferred location on the screen. This alternative -- motor noise -- account, however, would not predict the effect of target centrality and the u-shaped relation between target location and precision because motor noise should lead to normally distributed touches around the inferred location. Thus, we take the results as support for the idea of a universal process that is well-approximated by the model. This cognitive process might be rooted in humans’ evolved cognitive architecture, which is later refined during ontogeny. The phylogenetic roots of these processes might possibly lie much deeper as primates from a wide range of species follow gaze [@rosati2009looking; @tomasello1998five; @itakura2004gaze; @kano2014cross]. Yet, similarities in overt behavior do not imply the same underlying cognitive processes. The present study defines clear performance signatures that can be explored in other species to test such evolutionary hypotheses.

An unexpected result was that imprecision was higher on the left compared to the right side of the screen. One candidate explanation for this pattern might be the dominance -- despite variation -- of right-handedness across cultures [@papadatou2020human]. As of now, this is mere speculation and future research should investigate the origins of this pattern in more detail.

The study has further limitations. The fact that performance in the task was correlated with exposure to touchscreens might have overshadowed other sources of variation. However, we think it is in an important innovation that we were able to account for this effect. Most developmental cross-cultural studies do not even question the portability of their measurement instruments. Importantly, the key result that the processing signatures were evident in all communities, is immune to this finding. The potential that lies in the precise individual-level measurement that our task achieves was largely unexploited. As mentioned above, the questionnaire items only offered a very coarse picture into children's actual lived experiences. Future work could increase the resolution with which everyday experiences in children from diverse communities are recorded to compare the drivers behind social-cognitive development as we observe it. Recent work in the field of language acquisition has shown how technological innovations allowed for direct recording of social interactions across communities which can be used to close this explanatory gap [@bergelson2023everyday; @donnelly2021longitudinal].

In sum, our work pioneers an approach that introduces computational modeling and precise individual-level measurement to the cross-cultural study of cognitive development. This approach allowed us to identify potential universals in the human cognitive architecture rather than just overt behavior. As such, it can serve as a blueprint for future research on a broad spectrum of cognitive abilities. Finally, the study provides a much-needed empirical foundation for theories on the nature of the human mind. Children from diverse communities deploy similar cognitive processes in interpreting gaze, pointing to a universal foundation of basic social cognition.
